---
title: "DATA 607 Project 1"
author: "Gregg Maloy"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r load packages, include=FALSE}
library(dplyr)
library(tidyverse)
library(data.table)
library(tidyr)
library(readr)
library("ggplot2")
```
### Introduction
The purpose of this assignment was tidy and transform data using the tidyr and dplyr packages.  For the assignment, an untidy CSV was created and loaded into R. 

### Part 1: Load File and Inspection
Below the untidy CSV was loaded into R via the read_csv command and placed into a 
dataframe. There were numerous data quality issues which needed to be addressed before any
analysis could be conducted.  Some of these issues included: a row which consisted of
entirely NA values, cells with missing values, inappropriate column names, etc.
```{r load data}
Airline_Arrivals<- suppressMessages(read_csv     
                      ("https://raw.githubusercontent.com/goygoyummm/Data607_R/main/20230223_Data_607_Assignment_5.csv", show_col_types = FALSE)) 

DF <- data.table(Airline_Arrivals)
DF
```
\newpage
### Part 2: Tidying the data
The following changes were made to tidy the data.  
1. Columns names were renamed where appropriate.    
2. Rows which did not contain any values were deleted.    
3. Cells which contained missing values were populated with appropriate values.    
4. A new variable was created from multiple variables to make the data more manageable   
and appropriate for analysis.  The variable 'Destination' replaced the variables 'Los     
Angelos','Phoenix','San Diego','San Francisco','Seattle'.    
5. The value 'on time' within the 'Arrival_Ontime_or_Delayed' variable was changed to    
'ontime' in order to delete a space.    
6. In a new dataframe, two new variables, 'ontime' and 'delayed', were created from the 
values within the 'Arrival_Ontime_or_Delayed' variable. 
```{r}
#rename columns
colnames(Airline_Arrivals)[1] ="Airline"
colnames(Airline_Arrivals)[2] ="Arrival_Ontime_or_Delayed"

#remove rows with all 'na' values
Airline_Arrivals <-Airline_Arrivals %>%
    filter_all(any_vars(!is.na(.)))

#fill missing values down
Airline_Arrivals<-Airline_Arrivals %>%
    fill(Airline, .direction = "down")

#transform to long format
Airline_Arrivals_Long = Airline_Arrivals %>%
  gather("Los Angelos", "Phoenix", "San Diego","San Francisco","Seattle", key =
           Destination, value = Count_of_Arrivals) %>%
          arrange(Airline,Arrival_Ontime_or_Delayed,desc(Count_of_Arrivals))

#remove spaces in values in one column
Airline_Arrivals_Long <-Airline_Arrivals_Long %>%
    mutate(Arrival_Ontime_or_Delayed = str_replace(Arrival_Ontime_or_Delayed, " ", ""))

#sort to make more intuitive
Airline_Arrivals_Long<-arrange(Airline_Arrivals_Long,Airline,Arrival_Ontime_or_Delayed,
                               desc(Count_of_Arrivals))

#Tranform to wide 
Airline_Arrivals_Wide <- Airline_Arrivals_Long %>% 
  spread(Arrival_Ontime_or_Delayed, Count_of_Arrivals)

#new column of total arrivals by airline by location
Airline_Arrivals_Wide <- Airline_Arrivals_Wide %>% 
rowwise() %>% 
  mutate(Total_Arrivals = sum(c_across(delayed:ontime), na.rm = F))
```
\newpage
### Part 2: Tidying the data continued
The result of the 'tidying' were two datasets, one in long format and the     
other in wide.     
    
Long Format
```{r}
DF1 <- data.table(Airline_Arrivals_Long)
DF1
```

Wide Format
```{r}
DF2 <- data.table(Airline_Arrivals_Wide)
DF2
```
\newpage
### Part 3: Calculations 
Below dplyr was utilized to perform for calculations for analysis.
```{r}
#All flights % of delays vs ontime 10,000 foot overview
  
Overview_All_Flights<- Airline_Arrivals_Long %>%
    dplyr::summarise(
        Airline='All_Airlines'
        , Delayed=
            sum(Count_of_Arrivals[Arrival_Ontime_or_Delayed == 'delayed'], na.rm=TRUE)
        , Total_Flights=sum(Count_of_Arrivals, na.rm=TRUE)
        , Percent_Ontime =(100-(Delayed/Total_Flights *100))
        , Percent_Delayed =(Delayed/Total_Flights)*100)

#All flights % of delays vs ontime drill down by airline
Overview_All_flights_by_Airline<- Airline_Arrivals_Wide %>%
    dplyr::group_by(Airline) %>% 
    dplyr::summarise(
                     Ontime=sum(ontime)
                    ,Delayed=sum(delayed)
                    ,Total_Arrivals= sum(Total_Arrivals) 
                    ,Percent_Ontime = 100 - (sum(delayed)/sum(Total_Arrivals)*100)
                    ,Percent_Delayed = sum(delayed)/sum(Total_Arrivals)*100)

#calculate two new columns for % arrivals delayed and % arrivals ontime by airline and destination 
Overview_All_flights_by_Airline_and_Destination<-Airline_Arrivals_Wide %>%
    dplyr::group_by(Destination,Airline,ontime,delayed,Total_Arrivals) %>% 
    dplyr::summarise(
                    ,Pct_Ontime = ((ontime/Total_Arrivals)*100)
                    ,Pct_Delayed = ((delayed/Total_Arrivals)*100))

#Rounding percentages for final tables
Overview_All_Flights<-Overview_All_Flights %>% 
  mutate(across(c(Percent_Ontime, Percent_Delayed), ~round(., 0)))

Overview_All_Flights_by_Airline<-Overview_All_flights_by_Airline %>% 
  mutate(across(c(Percent_Ontime, Percent_Delayed), ~round(., 0)))

Overview_All_Flights_by_Airline_and_Destination<-Overview_All_flights_by_Airline_and_Destination %>% 
mutate(across(c(Pct_Ontime, Pct_Delayed), ~round(., 0)))
```
\newpage
### Part 4: Analysis
First I took a look at the total number of arrivals which were ontime regardless of 
airline. 88% of all flights arrived on time.
```{r}
DT1 <- data.table(Overview_All_Flights)

# Overview_All_flights
ggplot(data=Overview_All_Flights, aes(x=Percent_Ontime , y=Airline)) +
  geom_bar(stat="identity", color="black", fill="Gray", width=0.25) +
  geom_text(aes(label=Percent_Ontime), vjust=-0.3, size=3.5) +
  theme_classic()
DT1
```  
\newpage
### Part 4: Analysis Continued   
Next I drilled down on the number of flights which were ontime, stratifying by airline.
89% of AM West airlines arrivals were ontime, while 87% of Alaska airlines arrivals were
ontime. Worth noting is AM West airline had ~3,000 more total arrivals than Alaska
airline.
```{r}
DT2 <- data.table(Overview_All_Flights_by_Airline)

ggplot(data=Overview_All_Flights_by_Airline, aes(x= Percent_Ontime, y=Airline,fill=Airline)) +
  geom_bar(stat="identity",color="black", position=position_dodge()) +
  #geom_bar(stat="identity", width=0.25) +
  geom_text(aes(label=Percent_Ontime), vjust=-0.3, size=3.5) +
  scale_fill_brewer(palette="Paired")+
  theme_classic()
DT2
``` 
\newpage
### Part 4: Analysis Continued 
Finally I drilled down by both airline and arrival destination.  Upon inspection of the
results it became apparent that Alaska airlines had a higher percentage of flights arrive
ontime for each destination.
```{r}
DT3 <- data.table(Overview_All_Flights_by_Airline_and_Destination)
ggplot(data=Overview_All_Flights_by_Airline_and_Destination, 
       aes(x= Pct_Ontime, y=reorder(Destination,+Pct_Ontime), fill=Airline))+
       geom_bar(stat="identity", color="black",position=position_dodge(width = 0.5))+
       geom_text(aes(label=Pct_Ontime), vjust=-0.3, position = position_dodge(0.9),
                 size=3.5)+
       scale_fill_brewer(palette="Paired", breaks=rev)+
  theme(axis.text.y=element_blank()) + #remove x axis labels
  theme_classic()   
DT3
```   
### Part 5: Conclusion
Tidyr and dplyr are very convenient tools for manipulating data, as well as performing analysis.

