---
title: "DATA 607 Project 2: NYC Public Transportation (3 of 3)"
author: "Gregg Maloy"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r load packages, include=FALSE}
library(dplyr)
library(tidyverse)
library(data.table)
#library(tidyr)
library(readr)
#library("ggplot2")

```
### Introduction
The goal of this assignment was to practice cleaning and manipulation of datasets for downstream
analysis work. For the assignment, an untidy CSV was obtained from the NYC.gov portal
(https://data.ny.gov/Transportation/MTA-Daily-Ridership-Data-Beginning-2020/vxuj-8kew) and loaded
into R. The dataset consist of NYC MTA daily ridership for time period 03/2020 - 03/2023. The
author of the blackboard post suggested the analysis 'could show how commuter travel varies based
on the day of the week compared to each line of transportation'.  

### Part 1: Load File and Inspection
Below the untidy CSV was loaded into R via the read.csv command and placed into a 
dataframe. There were numerous data quality issues which needed to be addressed before any
analysis could be conducted.  Some of these issues included: transformation of the dataset to
long format, deleting extraneous rows, renaming variables and converting some of the variables to date format.
```{r load data}
PT<- read.csv("https://raw.githubusercontent.com/goygoyummm/Data607_R/main/20230302_CUNY_DATA_607_Project_2_Public_Transportation.csv")

DF <- data.table(PT)
DF
```
\newpage
### Part 2: Tidying the data and calculations
The following changes were made to tidy the data.  
1. Dataframe transformation to long format    
2. Utilizing a case statement to create a new variable 'Type_of_Transportation'    
3. Formatting a variable as a date     
4. Creating a new variable 'Day' to provide the day of the week of ridership    
5. Creating new variables to find mean ridership for a particular day of the week     
6. Another transformation of the dataframe to a long format
```{r}
#transform to long format
DF1 = DF %>%
   gather("Subways..Total.Estimated.Ridership", "Buses..Total.Estimated.Ridership",
          "LIRR..Total.Estimated.Ridership","Metro.North..Total.Estimated.Ridership",
          "Access.A.Ride..Total.Scheduled.Trips","Bridges.and.Tunnels..Total.Traffic",
          "Staten.Island.Railway..Total.Estimated.Ridership", 
          key =Type_of_Public_Transportation , 
          value = Total_Estimated_Ridership )  %>%
   gather("Subways....of.Comparable.Pre.Pandemic.Day", "Buses....of.Comparable.Pre.Pandemic.Day",
          "LIRR....of.Comparable.Pre.Pandemic.Day",
          "Metro.North....of.Comparable.Pre.Pandemic.Day",
          "Access.A.Ride....of.Comparable.Pre.Pandemic.Day",
          "Bridges.and.Tunnels....of.Comparable.Pre.Pandemic.Day",
          "Staten.Island.Railway....of.Comparable.Pre.Pandemic.Day",
          key =Type_of_Public_Transportation_2, 
          value = Percent_of_Comparable_Pre_Pandemic_Day) %>%
          filter(Date==Date)

#pulling value for below case statement
rename<-unique(DF1$Type_of_Public_Transportation)
view(rename)

#case statement to make new variable
DF1 <- DF1%>% 
  mutate(Type_of_Transport = case_when(
    Type_of_Public_Transportation == 'Subways..Total.Estimated.Ridership'     ~ 'Subway',
    Type_of_Public_Transportation == 'Buses..Total.Estimated.Ridership'       ~ 'Bus',
    Type_of_Public_Transportation == 'LIRR..Total.Estimated.Ridership'        ~ 'LIRR',
    Type_of_Public_Transportation == 'Metro.North..Total.Estimated.Ridership' ~ 'MNRR',
    Type_of_Public_Transportation == 'Access.A.Ride..Total.Scheduled.Trips'   ~ 'AR',
    Type_of_Public_Transportation == 'Bridges.and.Tunnels..Total.Traffic'     ~ 'BT',
    Type_of_Public_Transportation == 'Staten.Island.Railway..Total.Estimated.Ridership' ~ 'SIRR'
))

DF2 <- subset(DF1, select = c(1,6,3,5))

DF2[['Date']] <- as.POSIXct(DF2[['Date']],
                                   format = "%m/%d/%Y")

DF2[['Day']]<- weekdays(DF2$Date)



DF3<- DF2 %>%
  dplyr::group_by(Type_of_Transport) %>% 
    dplyr::summarise(
 Monday_Average_Ridership = mean(Total_Estimated_Ridership[Day=='Monday'], na.rm = TRUE) 
,Tuesday_Average_Ridership = mean(Total_Estimated_Ridership[Day=='Tuesday'], na.rm = TRUE) 
,Wednesday_Average_Ridership = mean(Total_Estimated_Ridership[Day=='Wednesday'], na.rm = TRUE) 
,Thursday_Average_Ridership = mean(Total_Estimated_Ridership[Day=='Thursday'], na.rm = TRUE)
,Friday_Average_Ridership = mean(Total_Estimated_Ridership[Day=='Friday'], na.rm = TRUE) 
,Saturday_Average_Ridership = mean(Total_Estimated_Ridership[Day=='Saturday'], na.rm = TRUE) 
,Sunday_Average_Ridership = mean(Total_Estimated_Ridership[Day=='Sunday'], na.rm = TRUE)) 


DF4 = DF3 %>%
   gather("Monday_Average_Ridership", "Tuesday_Average_Ridership",
          "Wednesday_Average_Ridership","Thursday_Average_Ridership",
          "Friday_Average_Ridership","Saturday_Average_Ridership",
          "Sunday_Average_Ridership", 
          key =Day_junk , 
          value = Mean_Estimated_Ridership )

DF4 <- DF4%>% 
  mutate(Day = case_when(
    Day_junk == 'Monday_Average_Ridership'       ~ 'Monday',
    Day_junk == 'Tuesday_Average_Ridership'      ~ 'Tuesday',
    Day_junk == 'Wednesday_Average_Ridership'    ~ 'Wednesday',
    Day_junk == 'Thursday_Average_Ridership'     ~ 'Thursday',
    Day_junk == 'Friday_Average_Ridership'       ~ 'Friday',
    Day_junk == 'Saturday_Average_Ridership'     ~ 'Saturday',
    Day_junk == 'Sunday_Average_Ridership'       ~ 'Sunday'
))

  
DF4 <- subset(DF4, select = c(1,4,3))


DF4$Day <- factor(DF4$Day,
    levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

  

Railroad<- DF4[which(DF4$Type_of_Transport=='LIRR'| DF4$Type_of_Transport=='MNRR'| DF4$Type_of_Transport=='SIRR'),]

Subway_Bus_and_Bridge_and_Tunnels<- DF4[which(DF4$Type_of_Transport=='Bus'| DF4$Type_of_Transport=='Subway'| DF4$Type_of_Transport=='BT'),]

Access_A_Ride<- DF4[which( DF4$Type_of_Transport=='AR'),]

```
\newpage


\newpage
### Part 3: Analysis
Below the day of the week was plotted against mean estimated ridership for all modes of
transportation.  Generally, rideship decreased as the week headed into the weekend, noting modest
decreases on Friday, Saturday and Sunday for the subway, bus, and bridges and tunnels. Of note,
Sunday is the only day of the week where mean ridership for bridges and tunnels exceeded bus
ridership.   
```{r}
ggplot(DF4,                                      
       aes(x = Day,
           y = Mean_Estimated_Ridership,
           fill = Type_of_Transport)) +
  geom_bar(stat = "identity",position = "dodge")  
```
    
I decided to take a closer look the commuter rails and access-a-ride modes of transport since the y-axis of the above plot obscured any relationship.  Below commuter rail and the access-a-ride  ridership are plotted separately to assess day of week ridership utilization.  The pattern previously noted, decreased weekend utilization, also extended to commuter railroad and access-a-ride modes of transport. 

```{r}
ggplot(Railroad,                                      
       aes(x = Day,
           y = Mean_Estimated_Ridership,
           fill = Type_of_Transport)) +
  geom_bar(stat = "identity",position = "dodge")  


ggplot(Access_A_Ride,                                      
       aes(x = Day,
           y = Mean_Estimated_Ridership,
           fill = Type_of_Transport)) +
  geom_bar(stat = "identity",position = "dodge")  

```


### Part 4: Conclusion
Tidyr and dplyr are very convenient tools for manipulating data, as well as performing analysis.
Mean ridership for all modes of transport decreased starting Friday, hitting the lowest ridership on Sunday.  

