---
title: "DATA 607 Assignment Week 9"
author: "Gregg Maloy"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
```{r load packages, include=FALSE}
library(dplyr)
library(tidyverse)
library(data.table)
library(httr)
library("ggplot2")
library("stringr")
library(jsonlite)
library(mice)
library(kableExtra)
```
### Part 1: Introduction
The New York Times web site provides a rich set of APIs,    
as described here: https://developer.nytimes.com/apis    
You’ll need to start by signing up for an API key.  Your task is   
to choose one of the New York Times APIs, construct an interface    
in R to read in the JSON data, and transform it into an R DataFrame.    

For this assignment I wanted to use the nytimes API to:    
1. Trend by year how many times the word 'autism' appeared in the nytimes articles     
since it was first published.    
2. Determine when the word 'autism' became more frequently used.

\newpage
### Part 2: Connecting to API and Retrieve and Tidy Data
Since the nytimes started publication, the word autism has been mentioned in 
4432 articles.

```{r}
api_key       <- "gxwgXId5xA73ugWGSiR2A0BAWebVq0bo"
key_word      <- "autism" 
Autism_search <- paste0("http://api.nytimes.com/svc/search/v2/articlesearch.json?q=",key_word,"&api-key=",api_key)     
Content       <- fromJSON(Autism_search) 
Content$response$meta$hits    
```
I initially attempted to include all 4,432 articles in my analysis but     
the API threw an error at 200 responses.  I then decided to limit     
my search to the last 10 years.    

In recent years autism has been diagnosed with greater    
frequency.  The increased diagnosis rate is thought related to     
changes in the diagnosis criteria, as well as a more educated public.    
I wanted to pinpoint the year(s) when the term autism started to    
to be used with a similar frequency as it is used today. In the     
time period 01/01/2013-12/31/2022, autism appeared in 1920    
articles.

```{r}
api_key       <- "gxwgXId5xA73ugWGSiR2A0BAWebVq0bo"
key_word      <- "autism" 
begin_date    <- "20130101" 
end_date      <- "20221231" 
Autism_search_2000 <- paste0("http://api.nytimes.com/svc/search/v2/articlesearch.json?q=",key_word,"&begin_date=",begin_date,"&end_date=",end_date,"&facet_filter=true&api-key=",api_key , sep="")
Content_2000  <- fromJSON(Autism_search_2000) 
Content_2000 $response$meta$hits        
```
The below code determine 191 API requests would be needed to pull all 1920
articles.  
```{r}
PageNumber <- round((Content_2000$response$meta$hits[1] / 10)-1) 
PageNumber
#PageNumber<-100
```
Below a loop is utilized to pull the information on all 1920 articles since nytimes.com limits API returns to ten returns per page.
```{r}
data <- list()# empty df

for(i in 0:PageNumber 
){ 
  # extract the data and flatten it with json 
  jsonautismfile <- fromJSON(paste0(Autism_search_2000, "&page=", i), flatten = TRUE) %>% data.frame() 
  # message("Processing page ", i)
  # Store data in empty data set 
  data[[i+1]] <- jsonautismfile
  # sleeps for 6 seconds 
  Sys.sleep(6) 
}

All_Autism_Search <- rbind_pages(data) #place into dataframe

#include only relevant columns
df2 <- subset(All_Autism_Search,select=c('response.docs.section_name','response.docs.pub_date'))

#new variable creation
df2[, "year"]  <- format(as.Date(df2$response.docs.pub_date), format = "%Y")

#new variable creation
df3 <- df2%>% mutate(count =1)
#df subset
df4<-df3 %>% subset(select=c(3,4))
```

### Part 3: Analysis
In the timeframe 01/01/2013-12/31/2022, the word autism was used with    
fairly consistent annual frequency, as evident by the below line graph.
```{r}
df5<-df4 %>%
   dplyr::group_by(year) %>%
   dplyr::summarise(
           Number_of_Articles=sum(count)
          )

ggplot(data=df5, aes(x=year, y=Number_of_Articles, group=1, label=Number_of_Articles))+
  geom_line(color="blue")+
  geom_point()+
  geom_text(nudge_y = 10)

data.table(df5)
```
Since the word autism was mentioned at a fairly consistent frequency     
within the past decade, I decided to change my timeframe to    
determine when the word autism started to be used with its current    
frequency.

In the time period 01/01/1850-12/31/2009, the word autism was utilized     
in 1864 articles.
```{r}
api_key       <- "gxwgXId5xA73ugWGSiR2A0BAWebVq0bo"
key_word      <- "autism" 
#print        <- "1"
#section_name <-"front+page"
#print_section<-"("A", "1")"
begin_date2    <- "18500101" 
end_date2      <- "20091231" 
Autism_search_v2000 <- paste0("http://api.nytimes.com/svc/search/v2/articlesearch.json?q=",key_word,"&begin_date=",begin_date2,"&end_date=",end_date2,
                  #"&print_page:",print,
                  #"&section_name=",section_name,
                  #"&print_section:",print_section,
                  "&facet_filter=true&api-key=",api_key , sep="")
Content_v2000       <- fromJSON(Autism_search_v2000) 
Content_v2000 $response$meta$hit
```
The API would need to be called 185 seperate times.
```{r}
PageNumber2 <- round((Content_v2000$response$meta$hits[1] / 10)-1) 
PageNumber2
#PageNumber<-100
```

```{r}
data2 <- list()# empty df

for(i in 0:PageNumber2 
){ 
  # extract the data and flatten it with json 
  jsonautismfile2 <- fromJSON(paste0(Autism_search_v2000, "&page=", i), flatten = TRUE) %>% data.frame() 
  # message("Processing page ", i)
  # Store data in empty data set 
  data2[[i+1]] <- jsonautismfile2
  # sleeps for 6 seconds 
  Sys.sleep(6) 
}

All_Autism_Search2 <- rbind_pages(data2) #place into dataframe

#include only relevant columns
dfz <- subset(All_Autism_Search2,select=c('response.docs.section_name','response.docs.pub_date'))

#new variable creation
dfz[, "year"]  <- format(as.Date(dfz$response.docs.pub_date), format = "%Y")

#new variable creation
dfx <- dfz%>% mutate(count =1)
#df subset
dfw<-dfx %>% subset(select=c(3,4))
```
As evident in the below plot, the use of the word 'autism' in nytimes articles became 
much more popular in the early to mid 2000's.
```{r}
dfv<-dfw %>%
   dplyr::group_by(year) %>%
   dplyr::summarise(
           Number_of_Articles=sum(count)
          )
dfv %>%
ggplot(aes(x=year, y=Number_of_Articles, group=1, label=Number_of_Articles))+
  geom_line(color="blue")+
  geom_point()+
  geom_text(nudge_y = 10, size=2)+ 
  theme(axis.text.x = element_text(angle = 90))+ 
  #theme(axis.text = element_text(size = 5))
  theme(text = element_text(size = 6)) 
data.table(dfv)


#geom_text(aes(label=Percent_Ontime), vjust=-0.3, size=3.5) +
```

```
\newpage
### Part 5: Conclusion
Scapping API's ia powerful tool.  This work could serve as the starting point for an analysis on the term autism in popular media.